Visual Diagnostics of a Model Explainer -- Tools for the Assessment of LIME Explanations

### 2020-06-05 paper.Rnw
more edits

% Abstract and keywords
\abstract[Summary]{The importance of providing explanations for predictions made by black-box models has led to the development of model explainer methods such as LIME (local interpretable model-agnostic explanations) \citep{ribeiro:2016}. LIME uses a surrogate model to explain the relationship between predictor variables from a black-box model and the black-box model predictions in a local region around a prediction of interest. However, the quality of the resulting explanations relies on how well the explainer model captures the black-box model in a specified local region. Here we introduce four visual diagnostics to assess the quality of LIME explanations: (1) explanation scatterplot, (2) prediction comparison plot, (3) assessment metric plot, and (4) feature heatmap. We apply the visual diagnostics to a forensics bullet matching dataset to show examples where LIME explanations depend on the tuning parameters and the explainer model oversimplifies the black-box model. Our examples raise concerns about claims made about LIME \citep{ribeiro:2016} that are similar to other criticism found in the literature (\citet{alvarezmelis:2018}, \citet{laugel:2018}, and \citet{molnar:2019}).}
\keywords{LIME, black-box models, interpretability, diagnostics}

@@

As a result of the various ways LIME explanations can fail, it is important to assess LIME explanations. We suggest the use of visual diagnostics for assessment. In this paper, we lay out the set of claims about LIME made by \citet{ribeiro:2016} and propose four visualizations for the assessment of these claims: (1) \emph{explanation scatterplot}, (2) \emph{prediction comparison plot}, (3) \emph{feature heatmap}, (4) \emph{assessment metric plot}. While LIME has been implemented for image, tabular, and text data, we will only focus on tabular data. For additional simplicity, we will only discuss classification prediction models with a dichotomous response variable and continuous predictor variables. However, the proposed diagnostics may be extended to a wider range of situations.

@@

In this section, we introduce four visual diagnostic plots for the assessment of LIME. The plots focus on different levels of application of LIME (e.g. on explanation versus a set of explanations) to assess the LIME claims from different perspectives:

\begin{enumerate}
\item \emph{Explanation Scatterplot} (Section \ref{plot1}): Comparison of the explainer and complex models for an individual prediction of interest.
\item \emph{Prediction Comparison Plot} (Section \ref{plot2}): Comparison of explainer and complex model predictions for a set of predictions.
\item \emph{Feature Heatmap} (Section \ref{plot3}): Comparison of features selected by LIME across applications of LIME with different tuning parameters.k
\item \emph{Assessment Metric Plot}: Comparison of performance metrics for LIME across applications of LIME with different tuning parameters. 
\end{enumerate}

@@

\begin{figure*}[!tp]
<<sine-plot1, out.width = '\\textwidth', fig.width = 10, fig.height = 4, warning = FALSE>>=

@@

plot1 <- ggplot() +
  geom_point(data = sine_poi_perms,
             mapping = aes(x = x1,
                           y = x2,
                           color = rfpred,
                           size = weights),
             alpha = 0.8,
             shape = 20) +
  scale_size(range = c(0.1, 3)) +
  geom_rect(data = sine_poi_bounds %>% filter(feature == "x1"),
            mapping = aes(xmin = lower,
                          xmax = upper,
                          ymin = -Inf,
                          ymax = Inf),
            alpha = 0.25,
            fill = "grey90",
            color = "firebrick") +
  geom_rect(data = sine_poi_bounds %>% 
              filter(feature == "x2"),
            mapping = aes(xmin = -Inf,
                          xmax = Inf,
                          ymin = lower,
                          ymax = upper,
                          color = coef),
            alpha = 0.25,
            fill = "grey90",
            color = "steelblue") +
  geom_point(data = sine_poi %>%
               mutate(shape = "Prediction \nof Interest"),
             mapping = aes(x = x1, y = x2, shape = shape, fill = rfprob_blue),
             size = 7,
             alpha = 0.8) +
  scale_shape_manual(values = 23) +
  scale_fill_gradient2(low = sinecolor_red,
                        mid = '#f7f7f7',
                        high = sinecolor_blue,
                        midpoint = 0.5,
                        limits = c(0, 1)) +
  scale_color_gradient2(low = sinecolor_red,
                        mid = '#f7f7f7',
                        high = sinecolor_blue,
                        midpoint = 0.5,
                        limits = c(0, 1)) +
  theme_bw(base_family = "Times", base_size = 10) +
  labs(x = TeX("x'_1"),
       y = TeX("x'_2"),
       size = "Weight",
       shape = "",
       fill = "Random \nForest \nProbability \nfor 'blue'",
       color = "Random \nForest \nProbability \nfor 'blue'",
       title = "Complex and Explainer Model Comparison Plot") + 
  guides(shape = guide_legend(order = 1)) + 
  theme(aspect.ratio = 1)

plot_grid(sine_poi_exp, plot1, align = "h")
@
\caption{\kgc{Edit...} (right) The explanation plot depicts the estimated parameters from the explainer model. The lengths of the bars represent the absolute values of the coefficients, and the colors of the bars represent the coefficient values. Since the magnitude of the coefficient associated with $x_2$ is longer than that of $x_1$, the explanation suggests that $x_2$ has more of an effect on the random forest prediction of interest than $x_1$ does. The explainer model comparison plot shows the simulated data colored by the random forest probabilities and sized by the assigned LIME weights. The prediction of interest is shown as the diamond shaped point. The quantile bins associated with $x_1$ and $x_2$ are depicted by the solid lines. The color of the line indicates whether the explainer model supports a prediction of 'blue' (blue line) or supports a prediction of 'red' (red lines). This figure shows that the quantile bins are not flexible enough to capture the relationship between the random forest predictions and the $x_1$ and $x_2$ values.}
\label{fig:sine-plot-step3ab}
\end{figure*}

@@

\subsection{Explanation Scatterplots} \label{plot1}

\autoref{fig:sine-plot-poi} shows the visualization of the explanation produced by \emph{lime} for the prediction of interest. The figure shows that the random forest returned a probability of \Sexpr{sine_poi_lime_default$label_prob[1]} that the observation of interest belongs to category blue {(denoted as the ``label" in the figure). The colored bars indicate the features chosen by LIME and whether they support or contradict a random forest classification of blue. \kgc{Add explanation for bins in the plot.}

The explanation for the prediction of interest depicted in Figure \ref{fig:sine-plot-poi} can be interpreted as follows. A random forest classification of blue is supported by the prediction of interest having a value of $x_2$ that is greater than \Sexpr{str_split(sine_poi_lime_default$feature_desc[[1]], pattern = " ")[[1]][1]}, but the prediction of interest having a value of $x_1$ that is greater than \Sexpr{str_split(sine_poi_lime_default$feature_desc[[2]], pattern = " ")[[1]][1]} and less than or equal to \Sexpr{str_split(sine_poi_lime_default$feature_desc[[2]], pattern = " ")[[1]][5]} provides support against a classification of blue. Since the weight associated with $x_2$ has a larger magnitude than the weight associated with $x_1$, LIME explains that the $x_1$ and $x_2$ locations of the prediction of interest provide more support for a random forest classification of blue over red. This explanation agrees with the random forest probability of \Sexpr{sine_poi_lime_default$label_prob[1]} for blue, even though both classify the observation incorrectly.

\autoref{fig:sine-plot-poi} also includes an "explanation fit" value. This value corresponds to the deviance ratio from the R package \emph{glmnet} \citep{simon:2011} for a ridge regression model, or in other words, this is the $R^2$ value associated with the explainer model. In this case, it is \Sexpr{round(sine_poi_lime_default$model_r2[1], 2)} suggesting that the explainer model is not a good linear fit. However, it is commonly accepted that $R^2$ has limitations for accessing the quality of fit of a model \citep{sapra:2014}, and it should not be the only metric used for assessment of a model. Based on this metric or this plot, it is not yet possible to make an informed assessment of the trustworthiness of the explanation.

The plot on the right of Figure \ref{fig:sine-plot-step3ab} is the complex and explainer model comparison plot for the prediction of interest from the \data. This figure shows the relationship between the random forest model and the LIME explainer model. The random forest model is represented by the simulated values of $x'_2$ versus $x'_1$ colored by the random forest predictions. The explainer model is represented by the solid horizontal and vertical line identify the boundaries of the quantiles that contain the prediction of interest. The color of the lines represents whether the coefficient of the corresponding indicator variable in the explainer model supports a random forest prediction of 'blue' (blue lines) or not (red lines). Additionally, the size of the simulated data points represents the proximity weight used to fit the ridge regression explainer model. The prediction of interest is represented by the diamond shaped point.

@@

\subsection{Prediction Comparison Plot} \label{plot2}

<<sine-mid-explainer-pred>>=
# Determine the explainer model predictions with the
# random forest predictions are between 0.25 and 0.75
mid_expred <- sine_lime_default %>%
  group_by(case, sim_method) %>%
  ungroup() %>%
  filter(label_prob > 0.25, label_prob < 0.75) %>%
  select(label_prob, model_prediction)
@

\begin{figure}[!t]
<<sine-plot-compare-bins, warning = FALSE, out.width = '0.5\\textwidth', fig.width = 6.75, fig.height = 5.5>>=

# Compute the average explainer model predictions per bin
bin_mean_exp_pred <- sine_lime_default %>%
  select(case, model_prediction) %>%
  group_by(case) %>%
  slice(1) %>%
  ungroup() %>%
  full_join(sine_lime_default %>%
              select(case, feature, feature_desc) %>%
              spread(feature, feature_desc),
            by = "case") %>%
  group_by(x1, x2) %>%
  summarise(mean_exp_pred = mean(model_prediction)) %>%
  ungroup() %>%
  separate(x1, c("x1_other", "x1_upper"), sep = " <= ") %>%
  separate(x1_other, c("x1_almost_lower", "x1_other"), sep = "x1") %>%
  select(-x1_other) %>%
  separate(x1_almost_lower, c("x1_lower", "x1_other"), sep = " <") %>%
  select(-x1_other) %>%
  separate(x2, c("x2_other", "x2_upper"), sep = " <= ") %>%
  separate(x2_other, c("x2_almost_lower", "x2_other"), sep = "x2") %>%
  select(-x2_other) %>%
  separate(x2_almost_lower, c("x2_lower", "x2_other"), sep = " <") %>%
  select(-x2_other) %>%
  mutate(x1_lower = ifelse(x1_lower == "", "-Inf", x1_lower),
         x1_upper = ifelse(is.na(x1_upper), "Inf", x1_upper),
         x2_lower = ifelse(x2_lower == "", "-Inf", x2_lower),
         x2_upper = ifelse(is.na(x2_upper), "Inf", x2_upper)) %>%
  mutate_at(.vars = c("x1_lower", "x1_upper",
                      "x2_lower", "x2_upper"),
            .funs = as.numeric) %>%
  arrange(x1_lower)

# Compute the lower and upper limits of the explainer model predictions
pred_col_lower = min(bin_mean_exp_pred$mean_exp_pred)
pred_col_upper = max(bin_mean_exp_pred$mean_exp_pred)

# Plot of the average feature weights by bin
ggplot() +
  geom_rect(data = bin_mean_exp_pred,
            mapping = aes(xmin = x1_lower,
                          xmax = x1_upper,
                          ymin = x2_lower,
                          ymax = x2_upper,
                          fill = mean_exp_pred),
            alpha = 1) +
  geom_point(data = sine_lime_default %>%
               select(case, feature, feature_value,
                      label_prob, model_prediction) %>%
               spread(feature, feature_value),
             mapping =  aes(x = x1, y = x2, color = label_prob)) +
  geom_point(data = sine_lime_default %>%
               select(case, feature, feature_value,
                      label_prob, model_prediction) %>%
               spread(feature, feature_value) %>%
               mutate(rf1 = ifelse(label_prob > 0.5, TRUE, FALSE),
                      lime1 = ifelse(model_prediction > 0.5,
                                     TRUE, FALSE)) %>%
               mutate(diff = rf1 - lime1) %>%
               select(case, x1, x2, diff) %>%
               filter(diff != 0) %>%
               mutate(shape = "Random Forest \nMisclassified \nPredictions"),
             mapping =  aes(x = x1, y = x2, color = label_prob,
                            shape = shape),
             alpha = 1,
             size = 3,
             color = "black") +
  scale_color_gradient2(low = "firebrick",
                        mid = '#f7f7f7',
                        high = "steelblue",
                        midpoint = 0.5,
                        limits = c(pred_col_lower, pred_col_upper)) +
  scale_fill_gradient2(low = "firebrick",
                       mid = '#f7f7f7',
                       high = "steelblue",
                       midpoint = 0.5,
                       limits = c(pred_col_lower, pred_col_upper)) +
  scale_shape_manual(values = 1) +
  xlim(-10, 10) +
  ylim(-10, 10) +
  theme_bw(base_family = "Times", base_size = 14) +
  labs(x = TeX("$x_1$"),
       y = TeX("$x_2$"),
       shape = "",
       fill = "Average \nExplainer \nModel \nProbability \nfor 'blue'",
       color = "Random \nForest \nProbability \nfor 'blue'",
       title = "Quantile Grid Comparison of LIME Explainer \nModel Predictions to Random Forest \nPredictions") +
  guides(shape = guide_legend(order = 1),
         fill = guide_legend(order = 2, reverse = T),
         color = guide_legend(order = 3, reverse = T))

@
\caption{Plot comparing the LIME explainer models predictions to the random forest predictions in the context of the four $x_1$ and $x_2$ quantile bins. The intersection of the quantile bins are colored by the average of the test data explainer model probabilities located in the intersection. The points are the test set data and colored by the corresponding random forest probability. The observations misclassificed by the random forest are indicated by black open circles.}
\label{fig:sine-plot-compare-bins}
\end{figure}

\autoref{fig:sine-plot-compare-bins} provides a visualization of the comparison of the explainer model predictions to the random forest predictions in the context of the $x_1$ and $x_2$ quantile bins. The intersection of the bins are colored by the average explainer model prediction of the test set observation located in the region. The \data \ test set is plotted as the circles and colored by the corresponding random forest probability. The test points that the random forest misclassified are encompassed by a black open circles. The plot highlights that the explainer models are over-simplifying the random forest predictions on the diagonal (lower left to upper right) where the explainer model predictions are close to 0.5, but the random forest probabilities range from 0 to 1. It is also possible to see the locations where the explainer model is over or under estimating the random forest model predictions on the off diagonals.

\subsection{Feature Heatmap Diagnostic Plot} \label{plot3}

@@

\begin{figure}[!t]
<<sine-feature-heatmap, out.width = '0.5\\textwidth', fig.width = 4, fig.height = 4>>=
feature_heatmap(sine_lime_explain$explain,
                order_method = "PCA") +
  theme_bw(base_family = "Times", base_size = 10) +
  theme(legend.position = "bottom",
        axis.text.y = element_blank(),
        axis.ticks.y = element_blank()) +
  labs(y = "Case")
@
\caption{Feature heatmap of the various LIME implementations applied to the \data. The cases from the test set are plotted on the y-axis, and the implementation methods are included on the x-axis. The colors of the tiles indicate the feature selected by LIME for the corresponding case and implementation method. The top faceted row includes the first features selected by LIME, and the bottom faceted row includes the second features selected by LIME. The faceted columns separate the kernel density implementation method from the quantile bin implementations. The vertical striping seen in the plot indicates that LIME is not consistent in selecting features across implementation methods.}
\label{fig:sine-feature-heatmap}
\end{figure}

Explanations produced by LIME are likely affected by exact choice of tuning parameters. An example of this was shown by Figures \ref{fig:concept-plot-good} and \ref{fig:concept-plot-bad} where the method used to weight the observations influenced the explanation. As of the time of writing this manuscript, there are no recommendations or procedures provided for how to determine which method to use besides for the default settings in \emph{lime} \kgc{(confirm this)}. In order to compare the explanations produced by LIME using different tuning parameters, we view an overview of all the explanations in the feature heatmap.

We plot the features selected by LIME for all of the cases and tuning parameters in heatmaps separated by the position of importance assigned by LIME in a plot that will be referred to as a feature heatmap. That is, for LIME applied for $t$ sets of tuning parameters to $n$ cases to select the $f$ top features, create $f$ heatmaps (one for each of the positions of importance) with the cases on the y-axis, the tuning parameters on the x-axis, and the cells colored by the feature chosen for the corresponding case and tuning parameter. This plot will be used assess the locality of the explanations and to compare results from different input options.

Two hypothetical examples of feature heatmaps are included in \autoref{fig:concept-heatmap}. The plots were created with the assumption that LIME was applied to select the top feature out of $p=8$ features for $n=10$ cases with $t=5$ sets of tuning parameters. If LIME were producing local explanations, it would be expected that the top features selected by LIME would vary for predictions located in different regions of the feature space unless a feature or several features are the deterministic features throughout the entire feature space. Additionally, if the tuning parameters used when implementing LIME did not affect the feature selected by LIME, the selected feature would be the same across all implementations for a case.

@@

\autoref{fig:sine-feature-heatmap} shows the feature heatmap for the LIME implementations applied to the \data. The top row of the plot contains the features selected as the most important by LIME, and the bottom row contains the second most important feature selected by LIME. For the quantile bin implementations, the original features prior to the interpretability transformation are plotted  since it is obvious that different features would be selected when the sizes of the bins change from one implementation to the next. This figure shows a clear difference in the explanations produced by LIME for the kernel density and two quantile bin methods and the other quantile bin methods. For both the kernel density and two quantile bin implementations, LIME selects $x_1$ as the most important feature and $x_2$ as the second most important feature across all cases in the test set. There is more variability in the features selected by LIME for the three to six quantile bin implementations. For these implementations, there is a mix of horizontal and vertical stripes, which suggests that LIME is not consistently providing the sample explanation across implementation methods. It is also interesting, that the implementations of 3 and 5 bins leads to the selection of the random noise variable of $x_3$ as an important variable in many predictions, which should not be the case.

\subsection{Assessment Metric Plot} \label{plot4}

\begin{figure}[!t]
<<sine-metric-plot, out.width = '0.5\\textwidth', fig.width = 4.5, fig.height = 4, warning = FALSE>>=

@@

@
\caption{Example of a metric comparison plot using the \data. Each row of the plot corresponds to one of three metrics: average fidelity, average $R^2$, or MSEE. The tuning parameters from the different LIME implementations are plotted on the x-axis, and the metric values are shown on the y-axis. The points are colored by rank within a metric indicating best to worst (dark to light). The kernel density simulation methods performs well according to all three metrics.}
\label{fig:sine-metric-plot}
\end{figure}

The feature heatmap for the \data \ in the previous section showed an example of how LIME explanations are not always consistent across \hh{this sounds like R versus python implementation. It's not what you mean} implementation methods. As a result, it is necessary to choose \hh{tuning parameters} that produce reliable explanations. One way to do this, \sout{would be} to apply LIME using different implementations, and then compute assessment metrics to determine the optimal tuning parameters. We will discuss several metrics that could be used for this purpose and present a comparison of the metrics in a plot. \kgc{***Right now, I have not included any variability in these plots. I could add standard deviation bars or change these to box plot.}

\kgc{I have moved all notation to the appendix. I will explain the metrics here in words.}

\begin{itemize}
\item $R^2$: Once LIME has been applied to all observations of interest, an average $R^2$ can be computed from the explainer models within an implementation method. 
\item Fidelity metric: for measuring the faithfulness of the explainer model to the complex model. This is done by comparing the complex and simple model predictions for each of the values in the simulated dataset accounting for the assigned weight. 
\item MSEE: meant to measure the faithfulness of the explainer model to the complex model by comparing their predictions, but instead of computing a squared difference over all observations in the simulated data, we only consider the observation of interest and compute the average squared deviation over the observations in the test set for an implementation of LIME. 
\end{itemize}

In \autoref{fig:sine-metric-plot}, the average fidelity is lowest for the kernel density simulation method increases as the number of quantile bins increase. This would suggest that the LIME implementation using the kernel density estimate provides the best local approximation to the random forest model on average. Note that the points in the plot are colored such that darker indicates a better metric value and lighter indicates a worse metric value.

@@

\begin{figure}[!t]
\centering
\includegraphics[width=3.5in]{./figure-static/signatures.png}
\label{fig:signatures}
\caption{Bullet signatures extracted from two bullets fired through the same barrel. The two signatures come from the same barrel-land and therefore have very similar patterns. The features used in the random forest from \citet{hare:2016} are different metrics that measure the similarity between two such signatures.}
\end{figure}

@@

\begin{figure*}[!b]
<<bullet-plot-rfpcp, out.width = '\\textwidth', warning = FALSE>>=

>>>>

% Abstract and keywords
\abstract[Summary]{The importance of providing explanations for predictions made by black-box models has led to the development of model explainer methods such as LIME (local interpretable model-agnostic explanations) \citep{ribeiro:2016}. LIME uses a surrogate model to explain the relationship between predictor variables from a black-box model and the black-box model predictions in a local region around a prediction of interest. However, the quality of the resulting explanations relies on how well the explainer model captures the black-box model in a specified local region. Here we introduce three visual diagnostics to assess the quality of LIME explanations: (1) explanation scatterplot, (2) assessment metric plot, and (3) feature heatmap. We apply the visual diagnostics to a forensics bullet matching dataset to show examples where LIME explanations depend on the tuning parameters and the explainer model oversimplifies the black-box model. Our examples raise concerns about claims made about LIME \citep{ribeiro:2016} that are similar to other criticism found in the literature (\citet{alvarezmelis:2018}, \citet{laugel:2018}, and \citet{molnar:2019}).}
\keywords{LIME, black-box models, interpretability, diagnostics}

@@

As a result of the various ways LIME explanations can fail, it is important to assess LIME explanations. We suggest the use of visual diagnostics for assessment. In this paper, we lay out the set of claims about LIME made by \citet{ribeiro:2016} and propose three visualizations for the assessment of these claims: (1) \emph{explanation scatterplot}, (2) \emph{feature heatmap}, (3) \emph{assessment metric plot}. While LIME has been implemented for image, tabular, and text data, we will only focus on tabular data. For additional simplicity, we will only discuss classification prediction models with a dichotomous response variable and continuous predictor variables. However, the proposed diagnostics may be extended to a wider range of situations.

@@

In this section, we introduce three visual diagnostic plots for the assessment of LIME. The plots focus on different levels of application of LIME (e.g. on explanation versus a set of explanations) to assess the LIME claims from different perspectives:

\begin{enumerate}
\item \emph{Explanation Scatterplot} (Section \ref{exp-scatter}): Comparison of the explainer and complex models for an individual prediction of interest.
\item \emph{Feature Heatmap} (Section \ref{feat-heat}): Comparison of features selected by LIME across applications of LIME with different tuning parameters.
\item \emph{Assessment Metric Plot} (Section \ref{assess-metric}): Comparison of performance metrics for LIME across applications of LIME with different tuning parameters. 
\end{enumerate}

@@

\begin{figure*}[!tp]
<<sine-exp-scatter, out.width = '\\textwidth', fig.width = 10, fig.height = 4, warning = FALSE>>=

@@

exp_scatter <- ggplot() +
  geom_point(data = sine_poi_perms,
             mapping = aes(x = x1,
                           y = x2,
                           color = rfpred,
                           size = weights),
             alpha = 0.8,
             shape = 20) +
  scale_size(range = c(0.1, 3)) +
  geom_rect(data = sine_poi_bounds %>% filter(feature == "x1"),
            mapping = aes(xmin = lower,
                          xmax = upper,
                          ymin = -Inf,
                          ymax = Inf),
            alpha = 0.25,
            fill = "grey90",
            color = "firebrick") +
  geom_rect(data = sine_poi_bounds %>% 
              filter(feature == "x2"),
            mapping = aes(xmin = -Inf,
                          xmax = Inf,
                          ymin = lower,
                          ymax = upper,
                          color = coef),
            alpha = 0.25,
            fill = "grey90",
            color = "steelblue") +
  geom_point(data = sine_poi %>%
               mutate(shape = "Prediction \nof Interest"),
             mapping = aes(x = x1, y = x2, shape = shape, fill = rfprob_blue),
             size = 7,
             alpha = 0.8) +
  scale_shape_manual(values = 23) +
  scale_fill_gradient2(low = sinecolor_red,
                        mid = '#f7f7f7',
                        high = sinecolor_blue,
                        midpoint = 0.5,
                        limits = c(0, 1)) +
  scale_color_gradient2(low = sinecolor_red,
                        mid = '#f7f7f7',
                        high = sinecolor_blue,
                        midpoint = 0.5,
                        limits = c(0, 1)) +
  theme_bw(base_family = "Times", base_size = 10) +
  labs(x = TeX("x'_1"),
       y = TeX("x'_2"),
       size = "Weight",
       shape = "",
       fill = "Random \nForest \nProbability \nfor 'blue'",
       color = "Random \nForest \nProbability \nfor 'blue'",
       title = "Complex and Explainer Model Comparison Plot") + 
  guides(shape = guide_legend(order = 1)) + 
  theme(aspect.ratio = 1)

plot_grid(sine_poi_exp, exp_scatter, align = "h")
@
\caption{(left) Visualization from the \emph{lime} R package of the LIME explanation for the \data individual prediction of interest. The bars represent the explainer model coefficients. (right) \emph{Explanation Scatterplot} The points represent the simulated data colored by the random forest probabilities and sized by the assigned LIME weights. The prediction of interest is shown as the diamond shaped point. The explainer model indicator variables associated with $x_1$ and $x_2$ (quantile bins containing the prediction of interest) are depicted by the solid lines. The color of the lines indicates the explainer model explanation for whether the corresponding feature value supports a prediction of 'blue' (blue lines) or supports a prediction of 'red' (red lines). This figure shows that the quantile bins are not flexible enough to capture the relationship between the random forest predictions and the $x_1$ and $x_2$ values.}
\label{fig:sine-exp-scatter}
\end{figure*}

@@

\subsection{Explanation Scatterplots} \label{exp-scatter}

The \emph{explanation scatterplot} is a visual diagnostic for assessing the LIME claims for an individual explanation by juxtaposing the complex and explainer models in one plot. The format of an explanation scatterplot depends on the LIME simulation method. We introduce the explanation scatterplot here under the default method in the \emph{lime} R package in which the data are simulated uniformly from four quantile bins. In this scenario, LIME converts continuous predictor variables to indicator variables identifying whether the variable value falls in the same quantile bin as the prediction of interest or not, and the indicator variables are used as the explainer model features. The explanation scatterplot achieves a juxtaposition of the complex and explainer models by plotting the LIME simulated data colored by the complex predictions overlaid by lines representing boundaries of the indicator variables used to fit the explainer model. The size of the points represents the weight assigned by LIME, and the color of the lines denote whether LIME indicates that a feature supports or contradicts a class prediction. A scatterplot is created for all pairwise combinations of features included in the LIME explanation. Appendix \ref{exp-scatter_plus} addresses the explanation scatterplot formats in other LIME simulation scenarios.

To demonstrate an explanation scatterplot, we focus on the misclassified point indicated in Figure \ref{fig:sine-plot-data} by the diamond as the prediction of interest. The left plot in Figure \ref{fig:sine-exp-scatter} shows the visualization of the explanation produced by \emph{lime} for the prediction of interest. The figure shows that the random forest returned a probability of \Sexpr{sine_poi_lime_default$label_prob[1]} that the observation of interest belongs to category blue {(denoted as the ``label" in the figure). The colored bars indicate the features chosen by LIME and whether they support or contradict a random forest classification of blue. Note that the features are represented as regions in the figure. These are the indicator variable regions created by LIME.

The explanation for the prediction of interest depicted in Figure \ref{fig:sine-exp-scatter} can be interpreted as follows. A random forest classification of blue is supported by the prediction of interest having a value of $x_2$ that is greater than \Sexpr{str_split(sine_poi_lime_default$feature_desc[[1]], pattern = " ")[[1]][1]}, but the prediction of interest having a value of $x_1$ that is greater than \Sexpr{str_split(sine_poi_lime_default$feature_desc[[2]], pattern = " ")[[1]][1]} and less than or equal to \Sexpr{str_split(sine_poi_lime_default$feature_desc[[2]], pattern = " ")[[1]][5]} provides support against a classification of blue. Since the weight associated with $x_2$ has a larger magnitude than the weight associated with $x_1$, LIME explains that the $x_1$ and $x_2$ locations of the prediction of interest provide more support for a random forest classification of blue over red. This explanation agrees with the random forest probability of \Sexpr{sine_poi_lime_default$label_prob[1]} for blue, even though both classify the observation incorrectly.

\autoref{fig:sine-exp-scatter} also includes an "explanation fit" value. This value corresponds to the deviance ratio from the R package \emph{glmnet} \citep{simon:2011} for a ridge regression model, or in other words, this is the $R^2$ value associated with the explainer model. In this case, it is \Sexpr{round(sine_poi_lime_default$model_r2[1], 2)} suggesting that the explainer model is not a good linear fit. However, it is commonly accepted that $R^2$ has limitations for accessing the quality of fit of a model \citep{sapra:2014}, and it should not be the only metric used for assessment of a model. Based on this metric or this plot, it is not yet possible to make an informed assessment of the trustworthiness of the explanation.

The plot on the right of Figure \ref{fig:sine-exp-scatter} is the complex and explainer model comparison plot for the prediction of interest from the \data. This figure shows the relationship between the random forest model and the LIME explainer model. The random forest model is represented by the simulated values of $x'_2$ versus $x'_1$ colored by the random forest predictions. The explainer model is represented by the solid horizontal and vertical line identify the boundaries of the quantiles that contain the prediction of interest. The color of the lines represents whether the coefficient of the corresponding indicator variable in the explainer model supports a random forest prediction of 'blue' (blue lines) or not (red lines). Additionally, the size of the simulated data points represents the proximity weight used to fit the ridge regression explainer model. The prediction of interest is represented by the diamond shaped point.

@@

The \data \ explanation only includes two features. In situations were more than two features are included in a LIME explanation, the explanation scatterplots can be extended to a generalized scatterplot matrix \kgc{(reference)} that includes all pairwise combinations of features. Generalized scatterplot matrices (and scatterplot matrices in general) fail to provide helpful information when the number of features becomes large \kgc{(reference)}. It is common in machine learning prediction problems for there to be a large number of features, and therefore, a generalized scatterplot matrix of explanation scatterplots would be ineffective. However, in the application of LIME, the user selects the number of features to return in the explanation (as previously mentioned). \citet{pedersen:2020} encourages users to select less than 10 features to include in the explanation \kgc{(need to check how to cite vignettes: https://cran.r-project.org/web/packages/lime/vignettes/Understanding\_lime.html)}. As a result, specifying a small number of features to include a LIME explanation leads to the feasibility of using generalized scatterplot matrices of explanation scatterplots to assess individual LIME explanations.

\subsection{Feature Heatmap} \label{feat-heat}

@@

\begin{figure}[!t]
<<sine-feat-heat, out.width = '0.5\\textwidth', fig.width = 4, fig.height = 4>>=
feature_heatmap(sine_lime_explain$explain,
                order_method = "PCA") +
  theme_bw(base_family = "Times", base_size = 10) +
  theme(legend.position = "bottom",
        axis.text.y = element_blank(),
        axis.ticks.y = element_blank()) +
  labs(y = "Case")
@
\caption{Feature heatmap of the various LIME implementations applied to the \data. The cases from the test set are plotted on the y-axis, and the implementation methods are included on the x-axis. The colors of the tiles indicate the feature selected by LIME for the corresponding case and implementation method. The top faceted row includes the first features selected by LIME, and the bottom faceted row includes the second features selected by LIME. The faceted columns separate the kernel density implementation method from the quantile bin implementations. The vertical striping seen in the plot indicates that LIME is not consistent in selecting features across implementation methods.}
\label{fig:sine-feat-heat}
\end{figure}

Explanations produced by LIME are likely to be affected by the choice of tuning parameters. An example of this was shown by Figures \ref{fig:concept-plot-good} and \ref{fig:concept-plot-bad} where the method used to weight the observations influenced the explanation. As of the time of writing this manuscript, there are no recommendations or procedures provided for how to determine which method to use besides for the default settings in \emph{lime} \kgc{(confirm this)}. In order to compare the explanations produced by LIME using different tuning parameters, we view an overview of all the explanations in the \emph{feature heatmap} visual diagnostic plot.

The feature heatmap uses colors to identify the features selected by LIME for sets of predictions and tuning parameters faceted by the position of importance assigned by LIME. That is, for LIME applied for $t$ sets of tuning parameters to $n$ cases to select the $f$ top features, create $f$ heatmaps (one for each of the positions of importance) with the cases on the y-axis, the tuning parameters on the x-axis, and the cells colored by the feature chosen for the corresponding case and tuning parameter. This plot is used assess the locality of the explanations and to compare results from different input options.

Two hypothetical examples of feature heatmaps are included in \autoref{fig:concept-heatmap}. The plots were created with the assumption that LIME was applied to select the top feature out of $p=8$ features for $n=10$ cases with $t=5$ sets of tuning parameters. If LIME produces local explanations, the top features selected by LIME will vary for predictions located in different regions of the feature space (unless a feature or several features are the deterministic features throughout the entire feature space). Additionally, if the tuning parameters did not affect the feature selected by LIME, the selected feature would be the same across all tuning parameters for a prediction.

@@

\autoref{fig:sine-feat-heat} shows the feature heatmap for the LIME implementations applied to the \data. The top row of the plot contains the features selected as the most important by LIME, and the bottom row contains the second most important feature selected by LIME. For the quantile bin implementations, the original features prior to the interpretability transformation are plotted  since it is obvious that different features would be selected when the sizes of the bins change from one implementation to the next. This figure shows a clear difference in the explanations produced by LIME for the kernel density and two quantile bin methods and the other quantile bin methods. For both the kernel density and two quantile bin implementations, LIME selects $x_1$ as the most important feature and $x_2$ as the second most important feature across all cases in the test set. There is more variability in the features selected by LIME for the three to six quantile bin implementations. For these implementations, there is a mix of horizontal and vertical stripes, which suggests that LIME is not consistently providing the sample explanation across implementation methods. It is also interesting, that the implementations of 3 and 5 bins leads to the selection of the random noise variable of $x_3$ as an important variable in many predictions, which should not be the case.

\subsection{Assessment Metric Plot} \label{assess-metric}

\begin{figure}[!t]
<<sine-assess-metric, out.width = '0.5\\textwidth', fig.width = 4.5, fig.height = 4, warning = FALSE>>=

@@

@
\caption{Example of a metric comparison plot using the \data. Each row of the plot corresponds to one of three metrics: average fidelity, average $R^2$, or MSEE. The tuning parameters from the different LIME implementations are plotted on the x-axis, and the metric values are shown on the y-axis. The points are colored by rank within a metric indicating best to worst (dark to light). The kernel density simulation methods performs well according to all three metrics.}
\label{fig:sine-assess-metric}
\end{figure}

The feature heatmap for the \data \ in the previous section showed an example inconsistent LIME explanations
across tuning parameter applications. As a result, it is necessary to choose tuning parameters that produce reliable explanations. One way to do this is to apply LIME using different implementations, and compute assessment metrics to determine the optimal tuning parameters. We will discuss three metrics that could be used for this purpose and present a comparison of the metrics in a plot.

Each metric presented below is computed on a set of LIME explanations associted with a set predictions of interest obtained using the the same tuning parameters.

\begin{itemize}
\item \emph{Average $R^2$}: A metric for the assessment of the model fit and linearity assumption computed as the average of ridge regression explainer models $R^2$ values (deviance ratios from the R package \emph{glmnet}).

\item \emph{Average Fidelity}: Provides comparison of the complex and simple model predictions for each of the values in the simulated dataset accounting for the assigned weight computed as the average of the fidelity metric presented in \citet{ribeiro:2016} (weighted distance between explainer and complex model predictions for simulated data).

\item \emph{Mean Squared Explanation Error (MSEE)}: Measures the faithfulness of the explainer model to the complex model by comparing their predictions similar to the average fidelity, but only the observation of interest is used to compute the average squared deviation over the observations in the test set.
\end{itemize}

Notation and formulas for these metrics are included in Appendix \ref{metric-details}.

\kgc{This is where I left off...}

\kgc{***Right now, I have not included any variability in these plots. I could add standard deviation bars or change these to box plot.}

In \autoref{fig:sine-assess-metric}, the average fidelity is lowest for the kernel density simulation method increases as the number of quantile bins increase. This would suggest that the LIME implementation using the kernel density estimate provides the best local approximation to the random forest model on average. Note that the points in the plot are colored such that darker indicates a better metric value and lighter indicates a worse metric value.

@@

@@

\begin{figure*}[!htp]
<<bullet-plot-rfpcp, out.width = '\\textwidth', warning = FALSE>>=
